---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Travis-CI Build Status](https://travis-ci.org/r-gris/tidync.svg?branch=master)](https://travis-ci.org/r-gris/tidync)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/r-gris/tidync?branch=master&svg=true)](https://ci.appveyor.com/project/r-gris/tidync) 
[![Coverage Status](https://img.shields.io/codecov/c/github/r-gris/tidync/master.svg)](https://codecov.io/github/r-gris/tidync?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/tidync)](https://cran.r-project.org/package=tidync)

# tidync

The goal of tidync is to ease exploring the contents of a NetCDF file and constructing efficient queries to extract
arbitrary hyperslabs. 

The data extracted can be used directly in array contexts, or in "long form" form "tidy" analysis and
visualization contexts. 


## Installation

You can install tidync from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("r-gris/tidync")
```

## Example

This is a basic example which shows you how to connect to a file. 

```{r}
file <- system.file("extdata", "unidata", "test_hgroups.nc", package = "tidync")
library(tidync)
tidync(file) 
```
See this article for more:  https://r-gris.github.io/tidync/articles/static-vignettes/tidync-examples.html

Stay tuned. 


## Code of conduct

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

