---
title: "Tidy NetCDF examples"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 10
    fig_height: 10
vignette: >
  %\VignetteIndexEntry{Tidy NetCDF examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



Find an arbitrary NetCDF file and print a traditional summary. 

```{r}
f <- system.file("extdata", "ifremer", "20130415.nc", package = "tidync")
library(RNetCDF)
print.nc(open.nc(f))
```

NetCDF is a very widely use file format for storing array-based data as *variables*. The **variable's** space is defined by its **dimensions** and their metadata. Dimensions are by definition "one-dimensional" consisting of one or more elements, a rectilinear virtual array with coordinate metadata on its units, type and interpretation. The **space** of a variable is defined as one or more of the dimensions in the file, but a variable won't necessarily use all the available dimensions and no dimensions are mandatory or particularly special. 

Some conventions exist to define usage and minimal standards for metadata for particular file schemas, but these are many, varied, and no particularly well adhered to in many contexts. 

A NetCDF file is essentially a container for simple array based data structures. There is limited capacity in the formal API for accessing data randomly within a variable, the primary mechanism is to define offset and stride (start and count) hyperslab indexes. (I.e. it's not possible to query a file for an arbitrary sparse set of values, without constructing a degenerate hyperslab query for each point or reading a hyperslab containing cells not in the set.)

Notice how the listing above is organized by *dimension* and then by *variable*. It's not particularly 
obvious that some variables live in the same set of dimensions as others. 

## tidync

Tidync provides facilities to explore the contents of a NetCDF file and construct efficient queries to extract arbitrary hyperslabs. These can be used directly in array contexts, or in "long form" database contexts. 

On first contact with the file, the available variables are reported (the first is made "active") and
the dimensions of the active variable are described.  The "active" variable may be specified with the `activate` function. 


```{r}
library(tidync)
tidync(f)

## activate another variable
tidync(f) %>% activate(concentration)
```

Here we see variables are clearly grouped by the *grid* they exist in, where grid is a specific (and ordered!) set of dimensions. This allows us to see the set of variables in a grid as existing in a single data set. We can expand out this entire set as a single data frame, which all the coordinates copied out - this is not efficient(!) but if we craft our queries sensibly to read only what we need, it's a very easy way to explore the data in a file. 

The 'hyper_filter' function allows specification of expressions to subset a variable based on each dimension's coordinate values.  

If no expressions are included we are presented with a table containing a row for each dimension, its extent in coordinates and its length. For convenience we also assign the activate form to an R variable, though we could just chain the entire operation without this. 

```{r}
concentration <- tidync(f) %>% activate(concentration) 

concentration %>% hyper_filter() 
```




By specifying inequality expressions we see an *implicit* subsetting of the array. Everything so far is implicit to 
delay any file-based computation required to actually interact with the file and read from it. 

Notice that these are "name = expr" paired expressions, because the right hand side may be quite general we 
need the left hand side name to be assured of the name of the dimension referred to. 

```{r}

concentration %>% hyper_filter(nj = nj < 20)


```

We can also use the special internal variable 'step', which will test against position in the dimension elements '1:length' rather than the values. It's not different in this case because ni and nj are just position dimensions anyway. The special 'dplyr' adverbs like 'between' will work. 

```{r}
concentration %>% hyper_filter(ni = index < 20, nj = dplyr::between(index, 30, 100))

```

## Data extraction

How to use these idioms to extract actual data? 

We can now exercise these variable choice and dimension filters to return actual data, either in by slicing out a  "slab" in array-form, or as a data frame. 

```{r}
hf <- concentration %>% hyper_filter(ni = index < 20, nj = dplyr::between(index, 30, 100))

## as an array
arr <- hf %>% hyper_slice()
str(arr)

## as a data frame

#concentration %>% hyper_tibble() %>% filter(!is.na(concentration))

```
