---
title: "NetCDF examples"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 10
    fig_height: 10
vignette: >
  %\VignetteIndexEntry{Tidy NetCDF examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of tidync is to ease exploring the contents of a NetCDF source and constructing efficient 
queries to extract arbitrary hyperslabs. The data extracted can be used directly as an array, or in 
"long form" form as a data frame for "tidy" analysis and visualization contexts. 

NetCDF is **Network Common Data Form** a very common, and very general way to store and work with 
scientific array-based data. NetCDF is defined and provided by [Unidata](https://www.unidata.ucar.edu/software/netcdf/).

Here we introduce traditional concepts of NetCDF, and show examples with built-in data to demonstrate tidync functionality. 

# NetCDF

NetCDF is a very widely used file format for storing array-based data as *variables*. The **variable's** space is defined by its **dimensions** and their metadata. Dimensions are by definition "one-dimensional" consisting of one or more elements, a rectilinear virtual array with coordinate metadata on its units, type and interpretation. The **space** of a variable is defined as one or more of the dimensions in the file, but a variable won't necessarily use all the available dimensions and no dimensions are mandatory or particularly special. 

Some conventions exist to define usage and minimal standards for metadata for particular file schemas, but these are many and varied, and not always adhered to. 

Both the RNetCDF and ncdf4 packages provide a traditional summary format, familiar to many NetCDF users as the output of the [command line program `ncdump`](https://www.unidata.ucar.edu/software/netcdf/netcdf-4/newdocs/netcdf/NetCDF-Utilities.html#NetCDF-Utilities). 


```{r}
f <- system.file("extdata", "ifremer", "20171002.nc", package = "tidync")
library(RNetCDF)
print.nc(open.nc(f))
```

The analogous operation at the command line on a suitable system. 

```bash 
ncdump -h /path/to/extdata/ifremer/20171002.nc
```



With the ncdf4 package it's a slightly different approach, but gives the same result. 

```R
print(ncdf4::nc_open(f))
```

Notice how the listing above is organized by *dimension* and then by *variable*. It's not particularly 
obvious that some variables live in the same set of dimensions as others. 

A NetCDF file is essentially a container for simple array based data structures. There is limited capacity in the formal API for accessing data randomly within a variable, the primary mechanism is to define offset and stride (start and count) hyperslab indexes. (I.e. it's not possible to query a file for an arbitrary sparse set of values, without constructing a degenerate hyperslab query for each point or reading a hyperslab containing cells not in the set.)



## tidync

Tidync aims to ease exploration of the contents of a NetCDF file and provides methods extract arbitrary hyperslabs. These can be used directly in array contexts, or in "long form" database contexts. 

On first contact with the file, the available variables are classified by grid and
dimension.  The "active" grid is the one that queries may be made against, and may be changed with the `activate` function. 


```{r}
library(tidync)
tidync(f)

```

Here we see variables are clearly grouped by the *grid* they exist in, where grid is a specific (and ordered!) set of dimensions. This allows us to see the set of variables that implicitly co-exist, they have the same *shape*.  The first grid "D0,D1,D2" has two variables, *concentration* and *quality_flag*, and the second "D2" has only one variable *time*. There are no general rules here, a file might have any number of dimensions and variables, and any variable might be defined by one or more dimensions. 

In this case the D2 grid has only one variable in its single dimension, and it happens to be a special kind of variable - a "coordinate dimension", as indicated by the `coord_dim` flag. In the traditional `ncdump` summary above it's easy to see there's only really one data grid, in `ni,nj,time` that it holds two variables, and that time is a special coordinate dimension - in contrast neither `ni` or `nj` have
an explicit 1-dimension variable. When there are many dimensions and/or many variables those patterns are *not* easy to see. 


We can expand out this entire set as a single data frame, which all the coordinates copied out - this is not efficient(!) but if we craft our queries sensibly to read only what we need, it's a very easy way to explore the data in a file. 

The 'hyper_filter' function allows specification of expressions to subset a variable based on each dimension's coordinate values.  

If no expressions are included we are presented with a table containing a row for each dimension, its extent in coordinates and its length. For convenience we also assign the activate form to an R variable, though we could just chain the entire operation without this. 

```{r}
concentration <- tidync(f) %>% activate(concentration) 

concentration %>% hyper_filter() 
```




By specifying inequality expressions we see an *implicit* subsetting of the array. Everything so far is implicit to 
delay any file-based computation required to actually interact with the file and read from it. 

Notice that these are "name = expr" paired expressions, because the right hand side may be quite general we 
need the left hand side name to be assured of the name of the dimension referred to. 

```{r}

concentration %>% hyper_filter(nj = nj < 20)


```

We can also use the special internal variable 'step', which will test against position in the dimension elements '1:length' rather than the values. It's not different in this case because ni and nj are just position dimensions anyway. The special 'dplyr' adverbs like 'between' will work. 

```{r}
concentration %>% hyper_filter(ni = index < 20, nj = dplyr::between(index, 30, 100))

```

## Data extraction

How to use these idioms to extract actual data? 

We can now exercise these variable choice and dimension filters to return actual data, either in by slicing out a  "slab" in array-form, or as a data frame. 

```{r}
hf <- concentration %>% hyper_filter(ni = index < 20, nj = dplyr::between(index, 30, 100))

## as an array
arr <- hf %>% hyper_slice()
str(arr)

## as a data frame

#concentration %>% hyper_tibble() %>% filter(!is.na(concentration))

```
